---
title: "defaulterPredictionCNN"
author: "Kanishk Kapoor"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}


install.packages("tensorflow")
install.packages("keras")

```

```{r}
library("dplyr")
library(tensorflow)
library(keras)

```


```{r}
l_data <- read.csv("bankloans.csv", stringsAsFactors = FALSE)
```

```{r}
#normalize numeric features

scale_data <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}


numerical_features <- c("age", "ed", "employ" , "address", "income", "debtinc", "creddebt", "othdebt")

```

```{r}

l_data[numerical_features] <- lapply(l_data[numerical_features], scale_data)

l_data$default <- as.factor(l_data$default)
```

In machine learning:
	•	Inputs (X) go into the model to make predictions.
	•	Outputs (y) are the true values used to calculate the error (loss).

Think of it like teaching:
	•	x_train: Homework questions
	•	y_train: Correct answers
	•	Model: Student trying to learn the pattern
	•	x_test: Exam questions
	•	y_test: Exam answer key (to grade performance)

```{r}
#TRAIN TEST SPLIT
l_data$default <- as.numeric(l_data$default == 1)

set.seed(123)
train_idx <- sample(1:nrow(l_data), 0.8 * nrow(l_data))
train_data <- l_data[train_idx, ]
test_data <- l_data[-train_idx, ]

# Convert to matrices for Keras
x_train <- as.matrix(train_data[, -9])  # Features
y_train <- as.matrix(as.numeric(train_data$default))  # Target
x_test <- as.matrix(test_data[, -9])
y_test <- as.matrix(as.numeric(test_data$default))

# Assuming x_train and y_train are matrices or data frames
valid_indices <- which(!is.na(y_train))
x_train_clean <- x_train[valid_indices, ]
y_train_clean <- y_train[valid_indices]


```
```{r}
valid_test_indices <- which(!is.na(y_test))
x_test_clean <- x_test[valid_test_indices, ]
y_test_clean <- y_test[valid_test_indices]
```



 Model Architecture (a Sequential Model):

model <- keras_model_sequential() %>%
  layer_dense(units = 16, activation = "relu", input_shape = c(ncol(x_train))) %>%
  layer_dense(units = 8, activation = "relu") %>%
  layer_dense(units = 1, activation = "sigmoid")

This is a 3-layer feedforward neural network, used for binary classification (predicting loan default or not).

⸻

 1. layer_dense(units = 16, activation = "relu", input_shape = c(ncol(x_train)))

 What’s happening here?
	•	This is the first hidden layer.
	•	Input: Each training example has multiple features (equal to the number of columns in x_train, like age, income, etc.).
	•	Neurons: 16 neurons in this layer.
	•	Each neuron: takes in the input features → performs a weighted sum → adds a bias → passes it through the ReLU activation function.

 ReLU Activation Function:

ReLU(x) = max(0, x)

	•	Removes negative values.
	•	Adds non-linearity — helps the model learn complex patterns like “high income but high debt still causes default”.

⸻

2. layer_dense(units = 8, activation = "relu")

 What’s happening here?
	•	This is the second hidden layer.
	•	It takes the output from the first hidden layer (16 values per sample), then:
	•	Multiplies them by weights
	•	Adds biases
	•	Applies ReLU again
	•	8 neurons = a further dimensionality reduction, learning even more abstract patterns from the previous layer.

💡 Think of this like moving from raw inputs → intermediate ideas → refined signals.

⸻

 3. layer_dense(units = 1, activation = "sigmoid")

 Final output layer:
	•	1 neuron → outputs a single value between 0 and 1.
	•	That’s because of the sigmoid activation:

Sigmoid(x) = 1 / (1 + e^(-x))

	•	This turns the model’s final prediction into a probability — how likely the user is to default on a loan.
	•	Closer to 1 → likely to default
	•	Closer to 0 → likely to repay

⸻

 In Summary:

Layer	Units	Activation	Purpose
Input Layer	–	–	Feeds data to the model (x_train)
Hidden Layer 1	16	ReLU	Learns basic features and interactions
Hidden Layer 2	8	ReLU	Learns deeper, refined patterns
Output Layer	1	Sigmoid	Predicts probability of default (0 to 1)



⸻

 How it works on x_train:

During training:
	1.	Each training example flows through all 3 layers (forward pass).
	2.	Output is compared to y_train (true labels).
	3.	The model calculates loss (error) and updates weights using backpropagation.


```{r}
#fnn MODEL


model <- keras_model_sequential() %>%
  layer_dense(units = 16, activation = "relu", input_shape = c(ncol(x_train_clean))) %>%
  layer_dense(units = 8, activation = "relu") %>%
  layer_dense(units = 1, activation = "sigmoid")

```

```{r}
summary(model)
```

```{r}


# Compile Model
model %>% compile(
  optimizer = optimizer_adam(),
  loss = loss_binary_crossentropy(),
  metrics = c("accuracy")
)

```
```{r}
sum(is.na(x_train_clean))  # should be 0
sum(is.infinite(x_train_clean))  # should be 0
sum(is.na(y_train_clean))  # should be 0
sum(is.infinite(y_train_clean))  # should be 0
```

```{r}
# Train Model
history <- model %>% fit(
  x_train_clean, y_train_clean,
  epochs = 50,
  batch_size = 10,
  validation_split = 0.2
)

```
```{r}
# Evaluate Model
model %>% evaluate(x_test_clean, y_test_clean)
```

An accuracy of ~84.9% on the test set is pretty solid — especially considering that training accuracy was ~83.7% and validation accuracy was ~83.2%. Your model is:
	•	Not overfitting — no huge gap between training, validation, and test accuracy.
	•	Generalizing well — it performs slightly better on the test set than expected!
	
```{r}
library(caret)

pred_probs <- model %>% predict(x_test_clean)
pred_class <- ifelse(pred_probs > 0.5, 1, 0)

confusionMatrix(
  factor(pred_class), 
  factor(y_test_clean),
  positive = "1"
)
```

```{r}
library(pROC)

roc_obj <- roc(y_test_clean, pred_probs)
plot(roc_obj, col = "blue", main = "ROC Curve")
auc(roc_obj)
```
AUC closer to 1 = better model. 0.85+ is very good!
```{r}
# Plot example
hist(pred_probs, breaks = 20, col = 'steelblue',
     main = "Distribution of Predicted Default Probabilities",
     xlab = "Predicted Probability of Default")
```
```{r}
# Convert matrix to data frame and assign column names
x_test_df <- as.data.frame(x_test_clean)
colnames(x_test_df) <- colnames(x_train)  # assuming same structure
```

```{r}
predicted_probs <- as.vector(pred_probs)
predicted_class <- ifelse(predicted_probs > 0.5, "Flag for Review", "Approve")

library(dplyr)
library(tibble)

predictions_table <- x_test_df %>%
  mutate(
    Actual_Default = y_test_clean,
    Predicted_Probability = round(predicted_probs, 3),
    Decision = predicted_class
  )
```

```{r}
# Show top 10 customers with highest default risk
predictions_table %>%
  arrange(desc(Predicted_Probability)) %>%
  head(10)
```

```{r}
library(ggplot2)

ggplot(predictions_table, aes(x = Predicted_Probability, fill = Decision)) +
  geom_histogram(binwidth = 0.05, color = "black", alpha = 0.7) +
  labs(title = "Predicted Default Probability Distribution",
       x = "Probability of Default", y = "Number of Customers") +
  theme_minimal()
```
```{r}

library(caret)  # for confusionMatrix
library(vip)    # optional, for plotting

# Baseline accuracy on test set
baseline <- model %>% evaluate(x_test_clean, y_test_clean)
baseline_acc <- baseline["accuracy"]

# Data frame to store drops in accuracy
importance <- data.frame(Feature = character(), Accuracy_Drop = numeric())

# Loop over each feature
for (i in 1:ncol(x_test_clean)) {
  x_temp <- x_test_clean
  x_temp[, i] <- sample(x_temp[, i])  # shuffle the i-th column
  
  # Predict with shuffled feature
  temp_acc <- model %>% evaluate(x_temp, y_test_clean, verbose = 0)
  acc_drop <- baseline_acc - temp_acc["accuracy"]
  
  importance <- rbind(importance, data.frame(
    Feature = colnames(x_test_clean)[i],
    Accuracy_Drop = acc_drop
  ))
}

# Sort by most important
importance <- importance %>% arrange(desc(Accuracy_Drop))
```

```{r}
# Plot
ggplot(importance, aes(x = reorder(Feature, Accuracy_Drop), y = Accuracy_Drop)) +
  geom_col(fill = "maroon") +
  coord_flip() +
  labs(title = "Feature Importance (Permutation Method)",
       x = "Feature", y = "Drop in Accuracy when Shuffled") +
  theme_minimal()
```

